Estimating differences in scale for high throughput sequencing analysis with ALDEx2

### 2023-08-21 outline.Rmd
third attempt

All HTS data are normalized after sequencing to make the different libraries comparable. There are two main normalization methods. The first type of normalization uses an internal standard. This can be as simple as converting the counts to proportions; i.e., by dividing the count for each part in a sample by the total count of the sample. Derivatives of this method include the TPM and RPKM approaches where the proportion is multiplied by a constant (TPM by 10^6) or by a pair of constants (RPKM by 10^6 and the length of the part).  The other internal-only normalization is the CLR, where the count of each part is divided by the geometric mean of the counts of the parts in each sample and a logarithm is taken of the resulting ratio. In this case, a small pseudo-count or prior is first applied. The second class of normalization uses an external standard. Normalizations such as the RLE, TMM and CSS methods assume that the majority of the parts are invariant and can be identified and then used to determine a relative scale between samples. Each sample is then normalized by multiplying or dividing by this relative scale to produce normalized counts.

It should be clear that all normalizations are ratios where the only difference is in how the denominator is chosen. A further complication arises during analysis when logarithmic transformations may or may not be applied, and the CLR (or other log-ratio transform) is the only transformation that explicitly uses as log-transformed data as the starting point. Several groups have pointed out that apparent differences between normalizations [2Paulson:2013aa, @Skinnider:2019vc] can be largely explained by differential application of logarithms [@Costea:2014aa, @erb:pc2020]. Indeed in many datasets a logarithm of one of the transforms exhibits behaviours that is similar to the logarithm of another transform [@GloorAJS2023].

>>>>

All HTS data are normalized after sequencing to make the different libraries comparable. There are two main normalization methods. The first type of normalization uses a standard that is internal to each sample. This can be as simple as converting the counts to proportions; i.e., by dividing the count for each part in a sample by the total count of the sample. Derivatives of this method include the TPM (transcripts per million) and RPKM (reads per kilobase per million [@Mortazavi:2008] approaches where the proportion is multiplied by several constants that depend on the instrument, gene characteristics in the sample and the sequencing depth.  The other internal-only normalization is the CLR (centred log-ratio) transform[@aitchison1982], where the count of each part is divided by the geometric mean of the counts of the parts in each sample and a logarithm is taken of the resulting ratio. In this case, a small pseudo-count or prior is first applied. The second class of normalization uses as a standard a measure that is determined externally to the sample being normalized. Normalizations such as the RLE (relative log expression)[@Anders:2010], TMM (trimmed mean of M values)[@Robinson:2010a] and CSS (cumulative sum scaling)[@Paulson:2013aa] methods assume that the majority of the parts are invariant and that they  can be identified and then used to determine a relative scale between samples. Each sample is then normalized by multiplying or dividing by this relative scale to produce normalized counts.

It should be clear that all normalizations are ratios where the only difference is in how the denominator is chosen. A further complication arises during analysis when logarithmic transformations may or may not be applied, and the CLR (or other log-ratio transform) is the only transformation that explicitly uses as log-transformed data as the starting point. Several groups have pointed out that apparent differences between normalizations [@Paulson:2013aa, @Skinnider:2019vc] can be largely explained by differential application of logarithms [@Costea:2014aa, @erb:pc2020]. Indeed in many datasets a logarithm of one of these transforms exhibits behaviours that is similar to the logarithm of another transform [@GloorAJS2023].
